---
title: "Intro to R: Data Manipulation and Exploration"
author: "Christina Maimone"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
params:
  notes: no
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
notes<-params$notes
```



# Dates

# Sorting


# Adding and Removing Columns


# Recoding/Replacing Values


# Aggregating 


# Apply Functions



# Setup

For the sections below, we'll use a built in data set, `mtcars`

```{r}
data(mtcars)
head(mtcars)
```


# Explore Data

## First Steps

We've covered some simple measures before, like `mean` and `summary`:

```{r}
mean(mtcars$mpg)
summary(mtcars$mpg)
```

And we've made frequency tables of values:

```{r}
table(mtcars$gear)
```

Although we can go beyond 1 variable to 2 or 3:

```{r}
table(mtcars$gear, mtcars$am) # rows, cols
```

`ftable` makes output for 3 dimension tables a bit easier to read

```{r}
ftable(mtcars$gear, mtcars$am, mtcars$cyl) # rows, subrows, columns
```

or proportion tables:

```{r}
prop.table(table(mtcars$gear))
```


And we've learned we can group our data and summarize by group too

```{r}
mtcars %>%
  group_by(am) %>%
  summarize(mean(mpg),
            mean(hp)) 
```

What else can we do?


# Statistics

## Correlation

Correlation between two variables

```{r}
cor(mtcars$mpg, mtcars$disp)
```

Or all pairwise comparisons:

```{r}
cor(mtcars)
```

Alternatively, covariance

```{r}
cov(mtcars)
```


## Testing Differences

### T-test

Do manual and automatic transmission cars have different mean miles per gallon?

```{r}
t.test(mtcars$mpg~mtcars$am)
```

Think of `~` as "as a function of".  So above, MPG as a function of transmission, meaning grouped by or explained by.

We don't have to use the formula syntax.  We can send data for two different groups:

```{r, eval=FALSE}
t.test(mtcars$mpg[mtcars$am==0], mtcars$mpg[mtcars$am==1])
```

In addition to the printout, you can get the various parts of the results

```{r}
t1 <- t.test(mtcars$mpg~mtcars$am)
names(t1)
```

```{r}
t1$statistic
```


### Regression

Let's predict MPG using a linear model.  First, a simple bivariate regression -- just one explanatory variable.

Basic syntax is 

```{r, eval=FALSE}
lm(y ~ x1 + x2 + x3, data=df)
```


```{r}
lm(mpg ~ hp, data=mtcars)
```

Default output isn't much.  You get a lot more with `summary`:

```{r}
r1 <- lm(mpg ~ hp, data=mtcars)
summary(r1)
```

Note that a constant (Intercept) term was added automatically.

As above, you can pull out the individual components as needed

```{r}
names(r1)
```

### Visualizing the result

```{r}
plot(mtcars$mpg ~ mtcars$hp)
abline(r1)
```

Label points with

```{r, eval=FALSE}
identify(mtcars$hp, mtcars$mpg, 
         labels=rownames(mtcars))
```


And visualizing diagnostics:

```{r}
plot(r1)
```


### Additional terms

How do we add more variables?  What about transforming variables?  Here is some of the formula syntax:

Symbol | Example | Description
-------|-----------|----------------------------------
`~` | `y ~ x1` | Defines the formula (necessary to create a formula object)
`+` | `y ~ x1 + x2` | Include the variable
`-` | `y ~ -1 +x1` | Delete a term, usually a 1 for the intercept
`:` | `y ~ x1 + x1:x2` | Interaction term 
`*` | `y ~ x1*x2` | Interaction between the variables and each individually; same as `y ~ x1 + x2 + x1:x2`
`^` | `y ~ (x1, x2, x3)^3` | Include variables and all interactions, up to 3-way interactions
`I()` | `y ~ I(x1^2)` | Wrapper for transforming variables without having to create a new variable
`poly()` | `y ~ poly(x1, 2)` | Creates polynomial terms up to the degree specified 



We can use `I()` to transform variables as part of the formula syntax.  We can add character or factor variables directly, and R will automatically create dummy (indicator) variables for us.

```{r}
r2 <- lm(mpg ~ hp + I(hp^2) + am + cyl, data=mtcars)
summary(r2)
```

We added cylinders as a numeric variable above, but maybe it should be categorical?  Also, a different way to specify the horsepower terms with `poly`

```{r}
r3 <- lm(mpg ~ poly(hp, 2, raw=TRUE) + am + I(factor(cyl)), data=mtcars)
summary(r3)
```

Note that it picked one group to be the base comparison group, and created variables for the other two.  If, instead of seeing offsets from a base group, we want indicator variables for all groups, we can drop the automatically added intercept term:

```{r}
r4 <- lm(mpg ~ -1 + poly(hp, 2, raw=TRUE) + am + I(factor(cyl)), data=mtcars)
summary(r4)
```

This is also useful if we need to constrain the fitted line to go through the origin.

We can plot fitted values for each variable:

```{r}
termplot(r4, rug=TRUE)
```

[More on `termplot`](http://www.clayford.net/statistics/tag/termplot/)

### Predicting values

If we want to predict new values from our fitted model, we can supply a data frame with column names matching the data we used to fit the model.  Let's see how changing hp affects the model.  Hold transmission and cylinders at constant values.

```{r}
newvals <- data.frame(hp=seq(50,250,10),
                      am=0,
                      cyl=4)
mpg_predict<- predict(r4, newvals)
mpg_predict
```

Note that I don't have to transform `hp` or `cyl` before calling `predict`.

```{r}
plot(mpg_predict ~ newvals$hp, 
     type="l", lwd=3, col="red", # line style
     main="am=0; cyl=4") # making note of other values
points(mtcars$mpg~mtcars$hp)
```


