---
title: "Exercises Part 2: Data Frames and Other Data Structures"
author: "Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
```


**Note: new exercises coming soon.**


# Exercise: Explore Differences in Selecting from `data.frame`

Make a data frame, called `mydata`, that has 3 variables: `var1` which has the numbers 1 to 10, `var2` which has the letters A-J, and `var3` which has any 10 numbers you want to include.

After you've made the data frame, then add a fourth column, `month`, with the first 10 months:

```{r}
mydata<-data.frame(var1=1:10, var2=LETTERS[1:10], 
                   var3=c(9,4.5,33,45.6,-14,3,7,0.2,0,7))
mydata$month <- month.name[1:10]
mydata
```

Run each command below to figure out what type of data you get back.  

Hint: Use the function `typeof()` to examine what is returned in each case.

```{r, eval=FALSE}
mydata[1]
mydata[[1]]
mydata$var2
mydata["var2"]
mydata[1, 1]
mydata[, 1]
mydata[1,]
mydata[-1,]
```

```{asis, echo=answers, eval=answers}
### Answer
```

```{asis, echo=answers, eval=answers}
Why is `mydata$var2` of type integer?  Factors are stored as integers.  Why is `var2` a factor in the first place?  Take a look at the help page for `data.frame` to see if you can find the option that turned letters into factors.   
```



# Exercise: Working with a `data.frame`

R has some built-n data sets.  One of them is called `iris`.  You can just use the `iris` object (it's a `data.frame`) without creating it first. 

Hint: if you want `iris` to show up in the Environment tab, load it into your environment with `data(iris)`.  Otherwise, you can still use it, but it may not show in that tab.

Get the dimensions of `iris`.  Then get a list of the names.

Output the first 10 rows of `iris`.

View the iris data frame in the RStudio data viewer.

Select from the `iris` data frame the observations where the `Sepal.Width` is less than 2.5.  Do the same, but add the condition that the `Sepal.Length` must also be less than 5.

Using `which.max`, select a row with maximum `Sepal.Length`.  Is there only one row with the maximum value? 

Rename the column `Petal.Width` to `petalwidth`.  Challenge: do this without hard coding in the column number.



```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
dim(iris)
names(iris)
head(iris, 10)
```

```{r, echo=answers, eval=FALSE}
View(iris) # or click on the name iris in the Environment tab to the right if you've loaded it there.
```

```{r, echo=answers, eval=answers}
iris[iris$Sepal.Width < 2.5,]
iris[iris$Sepal.Width < 2.5 & iris$Sepal.Length < 5,]
iris[which.max(iris$Sepal.Length),] # which.max only returns first index of a maximum value
iris[iris$Sepal.Length == max(iris$Sepal.Length),] # check for other max Sepal.Length rows
names(iris)[names(iris)=='Petal.Width'] <- 'pedalwidth'
names(iris)
```


# Exercise: Working with a `data.frame`: `mtcars`

Another built-in data set is called `mtcars`.

Explore `mtcars`.  Look at the help page for mtcars to see the variable definitions. 

* Get the cars with 6 cylinders
* Get just the weight (wt) column
* Make a new data frame called fuel_efficient cars that has cars with mpg > 25
* Do the fuel efficient cars have lower average horsepower (hp) than the overall average?
* Challenge: Which car with 3 gears has the highest horsepower (hp)?  Hint: subset first, save as new variable, then find max



```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
mtcars[mtcars$cyl==6,]
mtcars$wt # or mtcars[,"wt"]
fuel_efficient <- mtcars[mtcars$mpg > 25,]
mean(fuel_efficient$hp) < mean(mtcars$hp)
tmp <- mtcars[mtcars$gear==3,]
tmp[which.max(tmp$hp),]
```


# Challenge Exercise: Use `subset`

Look up the help for the `subset` function.  Can you use it to repeat some of the operations in the two exercises above?

Note: we'll cover this function later in the workshop.


# Exercise: Select elements from a list

The `repurrrsive` package has example list objects in it.  Load the package (install first if needed).  Then:

* View the `sw_people` list with the `View` function.  What's this a list of?
* Get the contents of first element in the list
* Get the name of the first person in the list
* What's the gender of the last person in the list?


```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=FALSE}
install.packages("repurrrsive")
```

```{r, echo=answers, eval=answers}
library(repurrrsive)
```

```{r, echo=answers, eval=FALSE}
View(sw_people)
```

```{r, echo=answers, eval=answers}
sw_people[[1]]
sw_people[[1]]$name
sw_people[[length(sw_people)]]$gender
```

# Exercise: Fix subsetting errors

A researcher is trying to select observations from the `iris` data frame for the setosa species, but gets an error:  

```{r, error=TRUE}
iris[Species="setosa"]
```

Correct this expression to select the observations the researcher wants.  Hint: there may be more than 1 thing wrong.

```{asis, echo=answers, eval=answers}
### Answer
```

```{asis, echo=answers, eval=answers}
Answer: Three mistakes, in the order that R will give you errors: 1) Use `==`, not `=`, to test for equality; 2) Include the name of the data frame before the column: `iris$Species`; and 3) The researcher forgot the comma after the condition to indicate to select all columns from the data frame.
```

```{r, echo=answers, eval=answers}
iris[iris$Species=="setosa",]
```


# Exercise: Counting with Conditionals

Using the built-in data set `mtcars`: 

* How many cars have `mpg` greater than 30?  
* How many have horse power (hp) less than 100?

Hint: you can use the `sum` function to count the number of `TRUE` observations in a vector (as we did with `sum(is.na(x))`).  This works because `as.numeric(TRUE) == 1` and `as.numeric(FALSE) == 0`.  

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, eval=answers, echo=answers}
sum(mtcars$mpg > 30)
sum(mtcars$hp < 100)
```


# Challenge Exercise: Matrix Manipulation

Using information from [Linear Algebra in R](https://www.math.uh.edu/~jmorgan/Math6397/day13/LinearAlgebraR-Handout.pdf) by Søren Højsgaard if needed, create a 10 x 5 matrix of random normal numbers and a 5 x 1 vector of values 1 to 5 and multiply them using matrix multiplication.  Transpose your result.

Create a 6 x 6 matrix of random normal draws and take it's inverse (solve it).  Extract the diagonal from the result.  

Hint: Generate random normal draws with `rnorm`: look it up to see the options.

Hint: Matrix multiplication operator is `%*%`.

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
X <- matrix(rnorm(50), nrow=10)
y <- 1:5
t(X %*% y) # t() is for transpose

X <- matrix(rnorm(36), nrow=6)
solve(X)
diag(solve(X))
```

```{asis, echo=answers, eval=answers}
Note that the specific numbers may differ when you run this code due to randomness.  To prevent numbers from changing, set the random seed with `set.seed(123)` where `123` is any integer.
```



# Challenge Exercise: List Indexing

Select the vector of letters from the following object.  

Do it both using names (hint: printing the object might give you a hint) and using index numbers.

Extra hint: use the object viewer in RStudio to view the list after you create it.  See if there's any functionality to help you with this challenge.

```{r}
nested_list <- list(level1 = list(level2 = list(letters = LETTERS)))
```

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
nested_list$level1$level2$letters
nested_list[[1]][[1]][[1]]
```


# Challenge Exercise: Use `apply`

Use the `apply` function to get the average value of each variable in `mtcars`.  You'll need to look at the help page to figure out how to use it.

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
apply(mtcars, 2, mean)
```




