---
title: "Exercises Part 3: Reading and Writing Data Files"
author: "Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: TRUE
---

```{r, echo=FALSE, eval=TRUE}
answers<-params$answers
```

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE, size = "small")
#knitr::opts_knit$set(root.dir=)
```


# Exercise: Read Data

Read in the data (a csv file) from https://goo.gl/dWrc9m to a data frame called `gapminder` (data is also in `data/gapminder5.csv` if you don't have an internet connection).  Do you need to make any adjustments to the defaults?  Note: you can specify a url instead of a filename/file path when using `read.csv`.

Look at the type of each column in the data you read in.  

View the observations for Belgium.

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=FALSE}
gapminder<-read.csv("https://goo.gl/dWrc9m") ## defaults OK here
str(gapminder)
View(gapminder[gapminder$country=="Belgium",])
```


```{asis, echo=answers, eval=answers}
Or, with `readr`:
```


```{r, echo=answers, eval=answers}
library(readr)
gapminder<-read_csv("https://goo.gl/dWrc9m")
```

# Exercise: Read Tab-delimited

Read in the tab-delimited file (states.txt) at https://goo.gl/AwnS4R (data is also at `data/states.txt`).  What changes do you need to make to read in tab-delimited data?  Are any other adjustments to the defaults warranted?

Change the column names to state, lat, and lon.  

Does anything look strange to you about this data?

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=answers}
statedata<-read.csv("https://goo.gl/AwnS4R", sep='\t', col.names=c("state", "lat","lon"))
dim(statedata)
```

```{asis, echo=answers, eval=answers}
There are more than 50 observations.  Take a look at the data, and you'll see some unfamiliar codes.  These correspond to US territories.  It's always good to check to make sure your data looks like you expect.

Above, with `readr`:
```

```{r, echo=answers, eval=answers}
library(readr)
statedata<-read_tsv("https://goo.gl/AwnS4R", col_names=c("state", "lat","lon"))
```

# Exercise: Write a data file

You read in the state data above.  Now write it to file as a CSV file, with sensible options.

```{asis, echo=answers, eval=answers}
### Answer
```

```{r, echo=answers, eval=FALSE}
# data read into a data frame called statedata above
write.csv(statedata, "statedatacopy.csv", row.names=FALSE)
```

```{asis, echo=answers, eval=answers}
Above, with `readr`:
```

```{r, echo=answers, eval=FALSE}
library(readr)
statedata<-write_csv(statedata, "statedatacopy.csv")
```


# Challenge Exercise: Reading Excel Files

Create an Excel file.  Read it into Excel with the `readxl` package.  How would you read in the second sheet?

```{asis, echo=answers, eval=answers}
### Answer
```

```{asis, echo=answers, eval=answers}
Create your own file.  Let's pretend it's called `mydata.xlsx`.  Then
```

```{r, echo=answers, eval=FALSE}
library(readxl)
mydata <- read_excel("mydata.xlsx")
```

```{asis, echo=answers, eval=answers}
To read in the second sheet, use the `sheet` argument to `read_excel`.
```



